A generalized quadratic estimate for random field nonstationarity \thanksref{code}

### 2016-03-10 paper/paper.tex
fixed a few typos

For readability it will be advantageous to briefly summarize the notational conventions used throughout this paper, some of which are borrowed from Cosmology and somewhat nonstandard in the statistics literature.
Variables taking values in $\Bbb R^d$ or $\Bbb R^{d\times d}$ will generally be written with bold font such as $\bs x, \bs y, \bs k,\bs \ell\in \Bbb R^d$ or $\bs A,\bs M \in \Bbb R^{d\times d}$. Indexing into vector or matrix coordinates are written with subscripts so that $\bs x_i\in\Bbb R$ denotes the $i^{\text{th}}$ coordinate of $\bs x\in\Bbb R^d$, for example. When $\bs z, \bs w\in\Bbb C^d$ we let the (non-Hermitian) dot product  be denoted by $\bs z \cdot \bs w = \bs z_1 \bs w_1 + \ldots + \bs z_d \bs w_d$.

Vector fields  $\bs \theta(\bs x):\Bbb R^d\rightarrow \Bbb R^d$ are also written bold so that \mbox{$\bs \theta(\bs x)=(\bs\theta_{1}(\bs x),\ldots, \bs\theta_{d}(\bs x))^T$} where $\bs \theta_i(\bs x):\Bbb R^d\rightarrow \Bbb R$. The Fourier transform of $\bs \theta(\bs x)$, for example, is applied coordinate-wise and written $\bs \theta_{\bs k}=(\bs \theta_{1,\bs k},\ldots, \bs \theta_{d,\bs k})^T$ where $\bs \theta_{j,\bs k}$ denotes the Fourier transform of $\bs \theta_j(\bs x)$ and is defined as
\[
 \bs\theta_{j,\bs k}= \int e^{-i\bs x\cdot \bs k} \bs \theta_{j}(\bs x) \frac{d\bs x}{(2\pi)^{d/2}}.
 \]
For a mean-zero stationary random field $\{Z(\bs x): \bs x\in\Bbb R^d\}$ the autocorrelation function is denoted  $C^Z(\bs x-\bs y) := E(Z(\bs x)Z(\bs y)^*)$ and the spectral density is denoted $C_{\bs\ell}^{ZZ} = {(2\pi)}^{d/2} C_{\bs\ell}^Z$ where, via our convention, $C_{\bs\ell}^Z$ denotes the Fourier transform of $C^Z(\bs x)$.
Parenthetical superscripts are reserved for indexing functions (and \textit{not} higher order derivatives). For example $f^{(1)}_{\bs k}, f^{(2)}_{\bs k}, f^{(3)}_{\bs k}, \ldots$ denotes a sequence of functions taking arguments $\bs k\in \Bbb R^d$ in the Fourier domain. This convention avoids indexing ambiguities and the subscript convention of the Fourier transform.

@@

The normalizing constant $A_{\bs \ell}$ is completely determined by the weights used on the terms $Z^{obs}_{\bs k+\bs \ell}Z^{obs}_{-\bs k}$ through the requirement that $\hat\phi_{\bs \ell}$ be unbiased up to first order in $\phi$, in particular, requiring that $E(\hat\phi_{\bs \ell}) = \phi_{\bs \ell} + \mathcal O(\phi^2)+ \mathcal O(\phi^3)+\cdots$. 
This results in the following an analytic characterization for $A_{\bs \ell}$
\begin{equation}
\label{in text eq of the normalizing constant}
A_{\bs \ell}
        \int
        \frac{\big|{\bs\xi_{\bs\ell}\!\cdot\! \bs C^{(1)}_{\bs k} - \bs\xi_{\bs\ell}\!\cdot\! \bs C_{\bs k+\bs \ell}^{(1)}}\big|^2}{C^{ZZobs}_{\bs k+\bs \ell}C^{ZZobs}_{\bs k}}\frac{d\bs k}{{(2\pi)}^{d/2}} = 1 + \mathcal O(\phi) + \mathcal O(\phi^2)+\cdots.
\end{equation}
Notice also that one is free to manually change weights used on each term $Z^{obs}_{\bs k+\bs \ell}Z^{obs}_{-\bs k}$ in  (\ref{explicit form of the qe in body of paper}). This may be advantageous for optimizing the sampling behavior of $\hat\phi_{\bs \ell}$ to specific applications. For example one may want to down-weight corrupted frequencies in a particular experimental setting. In this case, however, the form of the normalizing constant $A_{\bs \ell}$ given in (\ref{in text eq of the normalizing constant}) will need to be adjusted accordingly.

One of the advantages of the estimator $\hat\phi_{\bs \ell}$, defined in (\ref{explicit form of the qe in body of paper}), is that there exists a fast algorithm for computing $\hat\phi_{\bs \ell}$ at all frequencies $\bs \ell$ simultaneously by alternating pointwise operations in the Fourier domain and the pixel domain. Indeed (\ref{explicit form of the qe in body of paper}) is equivalent to
\begin{align}
    \label{fast form of the qe in body of paper}
    \hat\phi_{\bs \ell}
    &=   A_{\bs \ell}\sum_{p=1}^d \bs\xi^*_{p,\bs \ell} \int 
        e^{-i\bs x\cdot\bs \ell} \mathscr A(\bs x)\mathscr B_{p}(\bs x)\frac{d\bs x}{{(2\pi)}^{d/2}}
\end{align}
where $\mathscr A_{\bs \ell}:= Z^{obs}_{\bs \ell}/ C^{ZZobs}_{\bs \ell}$ and $\mathscr B_{p,\bs \ell} := i 2  \, \textrm{imag}(\bs C^{(1)}_{p,\bs \ell}) Z^{obs}_{\bs\ell} / C^{ZZobs}_{\bs\ell}$ which  can be computed in $\mathcal O(n\log(n))$ time (when observing $Z^{obs}$ on a grid of size $n$) by a sequence of fast Fourier transforms, inverse fast Fourier transforms and pointwise operations.

The derivation above only depends on the local invariance property insofar as it is used to optimize the weights  in (\ref{explicit form of the qe in body of paper}) and the resulting normalizing constant given in (\ref{fast form of the qe in body of paper}). Indeed, exactly similar arguments can be made for deriving quadratic estimates of nonstationary models which are not locally invariant, such as covariance functions of the form $C^{\bs\theta}(\bs x,\bs y)=K(\bs x-\bs y,\bs\theta(\bs x)+\bs\theta(\bs y))$ for example. The key difference is that the quadratic estimate $\hat \phi_{\bs \ell}$ for models which are {\it not} locally invariant  tend to either have a large  $\mathcal O(\phi^2)+ \mathcal O(\phi^3)+\cdots$ bias,  small signal to noise ratio, or have significant non-Gaussian estimation variability. Local invariant models, in contrast, encourage a significant amount of cancellation occurring within $\mathcal O(\phi^2)+ \mathcal O(\phi^3)+\cdots$ so that bias is small even in the regime of moderately large signal to noise ratio. Moreover, small higher order terms provide a regime where accurate prediction of mean square estimation variability is accurately approximated with easily computable formulas for second moments of estimation variability. This is explored in more detail in Section \ref{Section: var and bias} and in the simulation examples presented later.

@@

In this section we show that $\hat \phi_{\bs \ell}$, derived in the previous section, is a special case of the original quadratic estimate developed in \cite{hu2001mapping, hu2002mass} for Cosmic Microwave Background gravitational lensing. Start by letting $Z(\bs x) = T(\bs x + \nabla \phi(\bs x))$  denote the lensed Cosmic Microwave Background  and $\phi(\bs x)$ denote the projected gravitational potential in the $\Bbb R^2$ flat sky approximation. In the original derivation \cite{hu2001mapping, hu2002mass} a Taylor approximation is first applied to $T(\bs x+ \nabla \phi(\bs x))$ as follows
\begin{equation}
\label{HuOk approx 1}
Z(\bs x) \approx T(\bs x) +  \nabla T(\bs x)\cdot\nabla \phi(\bs x)
\end{equation}
The above linear model is then used to linearly approximate $Z(\bs x)Z(\bs y)$ by additionally discarding the term $\big(\nabla T(\bs x)\cdot\nabla \phi(\bs x)\big) \big(\nabla T(\bs y)\cdot\nabla \phi(\bs y)\big)$ which is quadratic in $\phi$. Taking Fourier transforms, then an expected value, results in the following approximation
\begin{equation}
\label{HuOk approx 2}
E(Z_{\bs k+\bs \ell}Z_{-\bs k}) \approx \frac{\phi_{\bs \ell}}{2\pi} \Big(\bs \ell\cdot(\bs k+\bs \ell)C^{TT}_{\bs k+\bs \ell} - \bs \ell\cdot\bs k C^{TT}_{\bs k}\Big).
\end{equation}
Notice that (\ref{HuOk approx 2}) is a special case of (\ref{eq: first order cov intro}), and hence a special case of $\hat\phi_{\bs \ell}$ in (\ref{explicit form of the qe in body of paper}), when setting $\bs \xi_{\bs \ell} = i\bs \ell$ and $\bs C^{(1)}_{\bs k} = \frac{i\bs k}{2\pi}  C^{TT}_{\bs k}$.

It is important to notice a particular subtlety when analyzing the accuracy of  (\ref{HuOk approx 2}) in terms of the magnitude of the discarded terms in the Taylor approximation (\ref{HuOk approx 1}). This subtlety can be illustrated by assuming the displacement $\nabla \phi(\bs x)$ is extremely large and happens to be a \textit{constant function of $\bs x$}. In this case one clearly has $Z(\bs x) \not\approx T(\bs x)$, i.e. the zeroth order Taylor approximation completely breaks down. Yet, in a distributional sense, the zeroth order Taylor approximation is perfect since $Z(\bs x)$ and $T(\bs x)$ have the same finite dimensional distributions (by the fact that $T(\bs x)$ is isotropic and $\nabla \phi(\bs x)$ is assumed to be constant).
Therefore one can not quantify the accuracy of (\ref{HuOk approx 2}) by a map level analysis of the individual discarded terms.
In fact we propose that (\ref{HuOk approx 2}) is accurate, not because the map level Taylor approximation (\ref{HuOk approx 1}) is good (for which it is not), but rather because $\nabla \phi(\bs x)$ is a locally invariant parameter and the corresponding nonstationary covariance
\[
C^{\bs \theta}(\bs x,\bs y)=E(Z(\bs x)Z(\bs y)) = C^T(\bs x - \bs y + \nabla\phi(\bs x) - \nabla\phi(\bs y))
\]
has an accurate Taylor approximation in $\nabla\phi(\bs x) - \nabla\phi(\bs y)$, vis-\`a-vis (\ref{first order term, intro}).

@@

\begin{definition}
\label{def: quad est applied to X}
For any function $X_{\bs k, \bs \ell}:\Bbb R^d\times \Bbb R^d \rightarrow \Bbb C$ let
$\hat\phi_{\bs \ell}\{X_{\bs k,\bs \ell}\}$ denote the quadratic estimate $\hat\phi_{\bs \ell}$ defined in (\ref{explicit form of the qe in body of paper}) but applied to the function $X_{\bs k, \bs \ell}$ rather than $Z^{obs}_{\bs k+\bs \ell}Z^{obs}_{-\bs k}$. In particular $\hat\phi_{\bs \ell}\{X_{\bs k,\bs \ell}\}$ is a function of $\bs \ell$  and satisfies
\[
\hat\phi_{\bs \ell}\{X_{\bs k,\bs \ell}\}
:=A_{\bs \ell}
        \int
        {\Big(\bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k} - \bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k+\bs \ell}\Big)}^{\! *}
        \frac{X_{\bs k,\bs \ell}}{C^{ZZobs}_{\bs k+\bs \ell}C^{ZZobs}_{\bs k}}\frac{d\bs k}{{(2\pi)}^{d/2}}.
\]
If, on the other hand, $X_{\bs k}$ and $Y_{\bs k}$ are both function of a single frequency argument $\bs k\in \Bbb R^d$ then  we define
\[
\hat\phi_{\bs \ell}\{X,\!Y\}
:=A_{\bs \ell}
        \int
        {\Big(\bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k} - \bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k+\bs \ell}\Big)}^{\! *}
        \frac{X_{\bs k+\bs \ell}Y_{-\bs k}}{C^{ZZobs}_{\bs k+\bs \ell}C^{ZZobs}_{\bs k}}\frac{d\bs k}{{(2\pi)}^{d/2}}.
\]
\end{definition}

@@

To derive $C_{\bs \ell}^{\text{var }\hat\phi}$ first let $X(\bs x)$ denote the mean zero stationary Gaussian random field which models the sum of the observational noise $N(\bs x)$ and the shape noise discussed in the previous paragraph. By propagating the random field $X$ through the quadratic estimate one obtains an estimate of variability of $\hat\phi_{\bs \ell}$ around its expected value. In particular
\begin{align}
E\big(\big[\hat\phi_{\bs \ell}-E(\hat\phi_{\bs \ell})\big]\big[\hat\phi_{\bs \ell^\prime}-E(\hat\phi_{\bs \ell^\prime})\big]^*\big) &\approx E\big(\hat\phi_{\bs \ell}\{X,\!X\} \, \hat\phi_{\bs \ell^\prime}\{X,\!X\}^*\big) =:  \delta^{\phantom{*}}_{\bs \ell-\bs \ell^\prime} C^{\text{var }\hat\phi}_{\bs \ell} \label{var aprox wick form}
\end{align}
where the existence of the spectral density $C^{\text{var }\hat\phi}_{\bs \ell}$ is guaranteed by the fact that $X(\bs x)$ is stationary so that $\hat\phi_{\bs \ell}\{X,X\}$ is stationary in the pixel domain (see Claim \ref{thm: quantify the var fluctuations of hat phi}).
Depending on which approximation one uses for the baseline stationary fluctuations in $Z(\bs x)$, the spectral density of $X(\bs x)$ can be defined in one of the following two ways
\begin{align}\label{Cxx def}
C^{X\!X}_{\bs \ell}=
\begin{cases}
C^{N\!N}_{\bs \ell} + {(2\pi)}^{d/2}C^{(0)}_{\bs \ell}, & \text{option 1;}\\
C^{N\!N}_{\bs \ell} + C^{ZZm}_{\bs \ell}, & \text{option 2.}
\end{cases}
\end{align}
Depending on what definition one uses for $C^{X\!X}_{\bs \ell}$, Claim \ref{thm: quantify the var fluctuations of hat phi} establishes that
\begin{align}
	\label{eq: spec den of hatphi on X}
	C^{\text{var }\hat\phi}_{\bs \ell} &=
        2 A_{\bs\ell}^2
        \int
        {\Big|\bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k} - \bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k+\bs \ell}\Big|}^{2}
        \frac{C^{X\!X}_{\bs k+\bs \ell}}{{(C^{ZZobs}_{\bs k+\bs \ell})}{}^2}
        \frac{C^{X\!X}_{\bs k}}{{(C^{ZZobs}_{\bs k})}{}^2}\frac{d\bs k}{{(2\pi)}^{d}}.
    \end{align}
 In certain situations the right hand side of (\ref{eq: spec den of hatphi on X}) can be simplified. Recall that in the definition of $\hat\phi_{\bs \ell}$, one has two options for defining $C^{ZZobs}_{\bs \ell}$, either $C^{N\!N}_{\bs \ell} + {(2\pi)}^{d/2}C^{(0)}_{\bs \ell}$ or $C^{N\!N}_{\bs \ell} + C^{ZZm}_{\bs \ell}$, depending on if one wants to use the prior spectral density $C^{\phi\phi}_{\bs \ell}$ for optimizing the quadratic estimate weights. If the choice of $C^{ZZobs}_{\bs \ell}$ matches the choice of  $C_{\bs \ell}^{X\!X}$ then one obtains the following simplification of (\ref{eq: spec den of hatphi on X})
\begin{align}
	\label{eq: spec den of hatphi on CZZmobs}
C^{\text{var }\hat\phi}_{\bs \ell} = 2{(2\pi)}^{-d/2} A_{\bs \ell}.
 \end{align}

@@

In this section we investigate the set of possible local spectral densities, for different values of the nonstationary function $\bs \theta(\bs x)$, under the nonstationary phase random field model of $Z(\bs x)$. Since the nonstationarity in $Z(\bs x)$ is exclusively due to local variation of a spectral phase, there is significant restriction on how local spectral densities vary with $\bs x$. For example, one can easily see that all the local spectral densities of $Z$ must have the same integral (so that the pointwise variance of $Z(\bs x)$ is a constant function of $\bs x$). In what follows we characterize further restrictions and, in doing so, identify a second source of estimation bias due to the curved nature of the local spectral models. The theory of optimal transport and the $L_2$-Wasserstein metric play an important role in the characterization of local stationary models. We refer the reader to the excellent book \cite{villani2003topics} for an overview of the subject.

>>>>

For readability it will be advantageous to briefly summarize the notational conventions used throughout this paper, some of which are borrowed from Cosmology and are somewhat nonstandard in the statistics literature.
Variables taking values in $\Bbb R^d$ or $\Bbb R^{d\times d}$ will generally be written with bold font such as $\bs x, \bs y, \bs k,\bs \ell\in \Bbb R^d$ or $\bs A,\bs M \in \Bbb R^{d\times d}$. Indexing into vector or matrix coordinates are written with subscripts so that $\bs x_i\in\Bbb R$ denotes the $i^{\text{th}}$ coordinate of $\bs x\in\Bbb R^d$, for example. When $\bs z, \bs w\in\Bbb C^d$ we let the (non-Hermitian) dot product  be denoted by $\bs z \cdot \bs w = \bs z_1 \bs w_1 + \ldots + \bs z_d \bs w_d$.

Vector fields  $\bs \theta(\bs x):\Bbb R^d\rightarrow \Bbb R^d$ are also written bold so that \mbox{$\bs \theta(\bs x)=(\bs\theta_{1}(\bs x),\ldots, \bs\theta_{d}(\bs x))^T$} where $\bs \theta_i(\bs x):\Bbb R^d\rightarrow \Bbb R$. The Fourier transform of $\bs \theta(\bs x)$, for example, is applied coordinate-wise and written $\bs \theta_{\bs k}=(\bs \theta_{1,\bs k},\ldots, \bs \theta_{d,\bs k})^T$ where $\bs \theta_{j,\bs k}$ denotes the Fourier transform of $\bs \theta_j(\bs x)$ and is defined as
\[
 \bs\theta_{j,\bs k}= \int e^{-i\bs x\cdot \bs k} \bs \theta_{j}(\bs x) \frac{d\bs x}{(2\pi)^{d/2}}.
 \]
For a mean-zero stationary random field $\{Z(\bs x): \bs x\in\Bbb R^d\}$ the autocorrelation function is denoted  $C^Z(\bs x-\bs y) := E(Z(\bs x)Z(\bs y)^*)$ and the spectral density is denoted $C_{\bs\ell}^{ZZ} = {(2\pi)}^{d/2} C_{\bs\ell}^Z$ where, via our convention, $C_{\bs\ell}^Z$ denotes the Fourier transform of $C^Z(\bs x)$.
Parenthetical superscripts are reserved for enumerating functions (and \textit{not} higher order derivatives). For example $f^{(1)}_{\bs k}, f^{(2)}_{\bs k}, f^{(3)}_{\bs k}, \ldots$ denotes a sequence of functions taking arguments $\bs k\in \Bbb R^d$ in the Fourier domain. This convention avoids indexing ambiguities and the subscript convention of the Fourier transform.

@@

The normalizing constant $A_{\bs \ell}$ is completely determined by the weights used on the terms $Z^{obs}_{\bs k+\bs \ell}Z^{obs}_{-\bs k}$ through the requirement that $\hat\phi_{\bs \ell}$ be unbiased up to first order in $\phi$, in particular, requiring that $E(\hat\phi_{\bs \ell}) = \phi_{\bs \ell} + \mathcal O(\phi^2)+ \mathcal O(\phi^3)+\cdots$ or equivalently that $\frac{1}{\phi_{\bs \ell}}E(\hat\phi_{\bs \ell}) = 1 + \mathcal O(\phi)+ \mathcal O(\phi^2)+\cdots$. 
This results in the following an analytic characterization for $A_{\bs \ell}$
\begin{equation}
\label{in text eq of the normalizing constant}
 A_{\bs \ell}
        \int
        \frac{\big|{\bs\xi_{\bs\ell}\!\cdot\! \bs C^{(1)}_{\bs k} - \bs\xi_{\bs\ell}\!\cdot\! \bs C_{\bs k+\bs \ell}^{(1)}}\big|^2}{C^{ZZobs}_{\bs k+\bs \ell}C^{ZZobs}_{\bs k}}\frac{d\bs k}{{(2\pi)}^{d/2}} = 1 + \mathcal O(\phi) + \mathcal O(\phi^2) + \cdots.
\end{equation}
Notice also that one is free to manually change weights used on each term $Z^{obs}_{\bs k+\bs \ell}Z^{obs}_{-\bs k}$ in  (\ref{explicit form of the qe in body of paper}). This may be advantageous for optimizing the sampling behavior of $\hat\phi_{\bs \ell}$ to specific applications. For example one may want to down-weight corrupted frequencies in a particular experimental setting. In this case, however, the form of the normalizing constant $A_{\bs \ell}$ given in (\ref{in text eq of the normalizing constant}) will need to be adjusted accordingly.

One of the advantages of the estimator $\hat\phi_{\bs \ell}$, defined in (\ref{explicit form of the qe in body of paper}), is that there exists a fast algorithm for computing $\hat\phi_{\bs \ell}$ at all frequencies $\bs \ell$ simultaneously by alternating pointwise operations in the Fourier domain and the pixel domain. Indeed by Claim \ref{appendix, claim: quad est derivations} of Appendix \ref{section: Detailed Proofs} the quadratic estimate given in (\ref{explicit form of the qe in body of paper}) is equivalent to
\begin{align}
    \label{fast form of the qe in body of paper}
    \hat\phi_{\bs \ell}
    &=   A_{\bs \ell}\sum_{p=1}^d \bs\xi^*_{p,\bs \ell} \int 
        e^{-i\bs x\cdot\bs \ell} \mathscr A(\bs x)\mathscr B_{p}(\bs x)\frac{d\bs x}{{(2\pi)}^{d/2}}
\end{align}
where $\mathscr A_{\bs \ell}:= Z^{obs}_{\bs \ell}/ C^{ZZobs}_{\bs \ell}$ and $\mathscr B_{p,\bs \ell} := i 2  \, \textrm{imag}(\bs C^{(1)}_{p,\bs \ell}) Z^{obs}_{\bs\ell} / C^{ZZobs}_{\bs\ell}$ which  can be computed in $\mathcal O(n\log(n))$ time (when observing $Z^{obs}$ on a grid of size $n$) by a sequence of fast Fourier transforms, inverse fast Fourier transforms and pointwise operations.

The derivation above only depends on the local invariance property insofar as it is used to optimize the weights  in (\ref{explicit form of the qe in body of paper}) and the resulting normalizing constant given in (\ref{fast form of the qe in body of paper}). Indeed, exactly similar arguments can be made for deriving quadratic estimates of nonstationary models which are not locally invariant, such as covariance functions of the form $C^{\bs\theta}(\bs x,\bs y)=K(\bs x-\bs y,\bs\theta(\bs x)+\bs\theta(\bs y))$ for example. The key difference is that the quadratic estimate $\hat \phi_{\bs \ell}$ for models which are {\it not} locally invariant  tend to either have a large  $\mathcal O(\phi^2)+ \mathcal O(\phi^3)+\cdots$ bias,  small signal to noise ratio, or have significant non-Gaussian estimation variability. Local invariant models, in contrast, encourage a significant amount of cancellation occurring within $\mathcal O(\phi^2)+ \mathcal O(\phi^3)+\cdots$ so that bias is small even in the regime of moderately large signal to noise ratio. Moreover, small higher order terms provide a regime where mean square estimation variability is accurately approximated with easily computable formulas. This is explored in more detail in Section \ref{Section: var and bias} and in the simulation examples presented later.

@@

In this section we show that $\hat \phi_{\bs \ell}$, derived in the previous section, is an extension of the original quadratic estimate developed in \cite{hu2001mapping, hu2002mass} for Cosmic Microwave Background gravitational lensing. Start by letting $Z(\bs x) = T(\bs x + \nabla \phi(\bs x))$  denote the lensed Cosmic Microwave Background  and $\phi(\bs x)$ denote the projected gravitational potential in the $\Bbb R^2$ flat sky approximation. In the original derivation \cite{hu2001mapping, hu2002mass} a Taylor approximation is first applied to $T(\bs x+ \nabla \phi(\bs x))$ as follows
\begin{equation}
\label{HuOk approx 1}
Z(\bs x) \approx T(\bs x) +  \nabla T(\bs x)\cdot\nabla \phi(\bs x)
\end{equation}
The above linear model is then used to linearly approximate $Z(\bs x)Z(\bs y)$ by additionally discarding the term $\big(\nabla T(\bs x)\cdot\nabla \phi(\bs x)\big) \big(\nabla T(\bs y)\cdot\nabla \phi(\bs y)\big)$ which is quadratic in $\phi$. Taking Fourier transforms, then an expected value, results in the following approximation
\begin{equation}
\label{HuOk approx 2}
E(Z_{\bs k+\bs \ell}Z_{-\bs k}) \approx \frac{\phi_{\bs \ell}}{2\pi} \Big(\bs \ell\cdot(\bs k+\bs \ell)C^{TT}_{\bs k+\bs \ell} - \bs \ell\cdot\bs k C^{TT}_{\bs k}\Big).
\end{equation}
Notice that (\ref{HuOk approx 2}) is a special case of (\ref{eq: first order cov intro}), and hence a special case of $\hat\phi_{\bs \ell}$ in (\ref{explicit form of the qe in body of paper}), when setting $\bs \xi_{\bs \ell} = i\bs \ell$ and $\bs C^{(1)}_{\bs k} = \frac{i\bs k}{2\pi}  C^{TT}_{\bs k}$.

It is important to notice a particular subtlety when analyzing the accuracy of  (\ref{HuOk approx 2}) in terms of the magnitude of the discarded terms in the Taylor approximation (\ref{HuOk approx 1}). This subtlety can be illustrated by assuming the displacement $\nabla \phi(\bs x)$ is extremely large and happens to be a \textit{constant function of $\bs x$}. In this case one clearly has $Z(\bs x) \not\approx T(\bs x)$, i.e. the zeroth order Taylor approximation completely breaks down. Yet, in a distributional sense, the zeroth order Taylor approximation is perfect since $Z(\bs x)$ and $T(\bs x)$ have the same finite dimensional distributions (by the fact that $T(\bs x)$ is isotropic and $\nabla \phi(\bs x)$ is assumed to be constant).
Therefore one can not quantify the accuracy of (\ref{HuOk approx 2}) by a map level analysis of the individual discarded terms in (\ref{HuOk approx 1}).
In fact we propose that (\ref{HuOk approx 2}) is accurate, not because the map level Taylor approximation (\ref{HuOk approx 1}) is good (for which it is not), but rather because $\nabla \phi(\bs x)$ is a locally invariant parameter and the corresponding nonstationary covariance
\[
C^{\bs \theta}(\bs x,\bs y)=E(Z(\bs x)Z(\bs y)) = C^T(\bs x - \bs y + \nabla\phi(\bs x) - \nabla\phi(\bs y))
\]
has an accurate Taylor approximation in $\nabla\phi(\bs x) - \nabla\phi(\bs y)$, vis-\`a-vis (\ref{first order term, intro}).

@@

\begin{definition}
\label{def: quad est applied to X}
For any function $X_{\bs k, \bs \ell}:\Bbb R^d\times \Bbb R^d \rightarrow \Bbb C$ let
$\hat\phi_{\bs \ell}\{X_{\bs k,\bs \ell}\}$ denote the quadratic estimate $\hat\phi_{\bs \ell}$ defined in (\ref{explicit form of the qe in body of paper}) but applied to the function $X_{\bs k, \bs \ell}$ rather than $Z^{obs}_{\bs k+\bs \ell}Z^{obs}_{-\bs k}$. In particular $\hat\phi_{\bs \ell}\{X_{\bs k,\bs \ell}\}$ is a function of $\bs \ell$  and satisfies
\[
\hat\phi_{\bs \ell}\{X_{\bs k,\bs \ell}\}
:=A_{\bs \ell}
        \int
        {\Big(\bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k} - \bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k+\bs \ell}\Big)}^{\! *}
        \frac{X_{\bs k,\bs \ell}}{C^{ZZobs}_{\bs k+\bs \ell}C^{ZZobs}_{\bs k}}\frac{d\bs k}{{(2\pi)}^{d/2}}.
\]
If, on the other hand, $X_{\bs k}$ and $Y_{\bs k}$ are both functions of a single frequency argument $\bs k\in \Bbb R^d$ then  we define
\[
\hat\phi_{\bs \ell}\{X,\!Y\}
:=A_{\bs \ell}
        \int
        {\Big(\bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k} - \bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k+\bs \ell}\Big)}^{\! *}
        \frac{X_{\bs k+\bs \ell}Y_{-\bs k}}{C^{ZZobs}_{\bs k+\bs \ell}C^{ZZobs}_{\bs k}}\frac{d\bs k}{{(2\pi)}^{d/2}}.
\]
\end{definition}

@@

To derive $C_{\bs \ell}^{\text{var }\hat\phi}$ first let $X(\bs x)$ denote the mean zero stationary Gaussian random field which models the sum of the observational noise $N(\bs x)$ and the shape noise discussed in the previous paragraph. By propagating the random field $X$ through the quadratic estimate one obtains an estimate of variability of $\hat\phi_{\bs \ell}$ around its expected value. In particular
\begin{align}
E\big(\big[\hat\phi_{\bs \ell}-E(\hat\phi_{\bs \ell})\big]\big[\hat\phi_{\bs \ell^\prime}-E(\hat\phi_{\bs \ell^\prime})\big]^*\big) &\approx E\big(\hat\phi_{\bs \ell}\{X,\!X\} \, \hat\phi_{\bs \ell^\prime}\{X,\!X\}^*\big) =:  \delta^{\phantom{*}}_{\bs \ell-\bs \ell^\prime} C^{\text{var }\hat\phi}_{\bs \ell} \label{var aprox wick form}
\end{align}
where the existence of the spectral density $C^{\text{var }\hat\phi}_{\bs \ell}$ is guaranteed by the fact that $X(\bs x)$ is stationary so that $\hat\phi_{\bs \ell}\{X,X\}$ is stationary in the pixel domain (see Claim \ref{thm: quantify the var fluctuations of hat phi}).
Depending on which approximation one uses for the baseline stationary fluctuations in $Z(\bs x)$, the spectral density of $X(\bs x)$ can be defined in one of the following two ways
\begin{align}\label{Cxx def}
C^{X\!X}_{\bs \ell}=
\begin{cases}
C^{N\!N}_{\bs \ell} + {(2\pi)}^{d/2}C^{(0)}_{\bs \ell}, & \text{option 1;}\\
C^{N\!N}_{\bs \ell} + C^{ZZm}_{\bs \ell}, & \text{option 2.}
\end{cases}
\end{align}
Now given $C^{X\!X}_{\bs \ell}$, Claim \ref{thm: quantify the var fluctuations of hat phi} establishes that
\begin{align}
	\label{eq: spec den of hatphi on X}
	C^{\text{var }\hat\phi}_{\bs \ell} &=
        2 A_{\bs\ell}^2
        \int
        {\Big|\bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k} - \bs\xi_{\bs \ell} \!\cdot\!\bs C^{{(1)}}_{\bs k+\bs \ell}\Big|}^{2}
        \frac{C^{X\!X}_{\bs k+\bs \ell}}{{(C^{ZZobs}_{\bs k+\bs \ell})}{}^2}
        \frac{C^{X\!X}_{\bs k}}{{(C^{ZZobs}_{\bs k})}{}^2}\frac{d\bs k}{{(2\pi)}^{d}}.
    \end{align}
 In certain situations the right hand side of (\ref{eq: spec den of hatphi on X}) can be simplified. Recall that in the definition of $\hat\phi_{\bs \ell}$, one has two options for defining $C^{ZZobs}_{\bs \ell}$, either $C^{N\!N}_{\bs \ell} + {(2\pi)}^{d/2}C^{(0)}_{\bs \ell}$ or $C^{N\!N}_{\bs \ell} + C^{ZZm}_{\bs \ell}$, depending on if one wants to use the prior spectral density $C^{\phi\phi}_{\bs \ell}$ for optimizing the quadratic estimate weights. If the choice of $C^{ZZobs}_{\bs \ell}$ matches the choice of  $C_{\bs \ell}^{X\!X}$ then one obtains the following simplification of (\ref{eq: spec den of hatphi on X})
\begin{align}
	\label{eq: spec den of hatphi on CZZmobs}
C^{\text{var }\hat\phi}_{\bs \ell} = 2{(2\pi)}^{-d/2} A_{\bs \ell}.
 \end{align}

@@

In this section we investigate the set of possible local spectral densities, for different values of the nonstationary function $\bs \theta(\bs x)$, under the nonstationary spectral phase random field model of $Z(\bs x)$. Since the nonstationarity in $Z(\bs x)$ is exclusively due to local variation of a spectral phase, there is significant restriction on how local spectral densities can vary with $\bs x$. For example, one can easily see that all the local spectral densities of $Z(\bs x)$ must have the same integral (so that the pointwise variance of $Z(\bs x)$ is a constant function of $\bs x$). In what follows we characterize further restrictions and, in doing so, identify a second source of estimation bias due to the curved nature of the local spectral models. The theory of optimal transport and the $L_2$-Wasserstein metric play an important role in the characterization of local stationary models. We refer the reader to the excellent book \cite{villani2003topics} for an overview of the subject.

The local invariance property of nonstationary spectral phase models implies that the local distribution of $Z(\bs x)$ is invariant to changes in the magnitude of $\bs \theta(\bs x)$. However, the local behavior of $Z(\bs x)$ \textit{is} sensitive to the gradient of $\bs \theta(\bs x)$. In particular, suppose $\bs\theta(\bs x)$ has the form $\bs A \bs x + \bs b$  where $\bs A\in \Bbb R^{d\times d}$ and $\bs b \in \Bbb R^d$. In this case the covariance function $C^{\bs \theta}(\bs x,\bs y)$ is still invariant to changes in $\bs b \in \Bbb R^d$ but is sensitive to changes in $\bs A\in \Bbb R^{d\times d}$. Indeed assuming $\bs\theta(\bs x)=\bs A \bs x + \bs b$ one has
\begin{align}
C^{\bs \theta}(\bs x,\bs y)
&=\int  \exp\!\big(i(\bs x-\bs y)\!\cdot\! (\bs k + \bs A^T\bs\eta_{\bs k})\big) \, C_{\bs k}\frac{d\bs k}{(2\pi)^d}
=\int\exp\!\big(i(\bs x-\bs y)\!\cdot\! \bs\omega\big)  \frac{d\lambda(\bs \omega)}{(2\pi)^d} \label{eq: local spectral density for nonstationary phase}
\end{align} 
where $\lambda$ is the spectral measure obtained by a change of variables $\bs\omega = \bs k + \bs A^T\bs\eta_{\bs k}$ (under appropriate measurability assumptions on $\bs \eta_{\bs k}$) Therefore when $\bs\theta(\bs x) = \bs A \bs x + \bs b$ in a local neighborhood about $\bs x$ the process $Z(\bs x)$ becomes locally stationary with local spectral measure given by $\lambda$.
